---
title: "Density Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CompStat)
```

# Introduction
This vignette details the usage of the density-estimation module of the 'CompStat' package.

The density estimation we implement here is based on kernel smoothers. The central functions are:

* *iter_bw_est* - Iterative bandwidth selection using AMISE.
* *compile_density* - Density estimation given a bandwidth and kernel.
* *plot_density* - Combination of the above into a simple call with graphical output.

In this vignette, we present our considerations for the development process of each of these components, although we will mainly be focusing on *iter_bw_est*.

# Function breakdown
## iter_bw_est
The *iter_bw_est* function is based on the algorithm:

1. Receive an initial bandwidth estimate $h_0$ or use the Silverman-rule for computing one.
2. For $m = 1, \ldots, M$ do the following:
  + Compute the best-estimate for the $L_2$-norm of the unknown density $f$
  using the formula (NRH Chapter 2, Unnumbered equation):
$$
\left|\left|\tilde{f}''_m\right|\right|_2^2 = \frac{1}{n^2h_m^6} \sum_{i=1}^{n}\sum_{j=1}^n \int H''\left(\frac{x-x_i}{h_m}\right) \cdot H''\left(\frac{x-x_j}{h_m}\right) dx
$$
  + Compute the AMISE-optimal bandwidth using the update rule (NRH Chapter 2, Equation 2.3)
$$
h_{m+1} = \left(\frac{||K||_2^2}{||\tilde{f}_m''||_2^2\sigma_K^4}\right)^{1/5}n^{-1/5}
$$
3. Return the sequence of bandwidths $(h_0, \ldots, h_M$).

The motivation for the algorithm is that (NRH Chapter 2, Equation 2.3) could be used to compute the optimal bandwidth if true $||f||_2^2$ is known. Since this is not the case, the above algorithm instead uses the current best-estimate for $||f||_2^2$ as a plug-in estimate. This leads to a sequence of optimal-given-the-previous bandwidths.

This leaves the question of whether the sequence of bandwidth obtained will converge to or even approximate the true, optimal bandwidth. See section XXX for an investigation of this issue.

## compile_density
The *compile_density* algorithm is a rather simple wrapper around the formula:
$$
\hat{f}(x) = \frac{1}{nh}\sum_{i=1}^n H\left(\frac{x-x_i}{h}\right)
$$
Hence it serves to simply compute the estimated density once a bandwidth has been selected. We present two different implementations:
* A simple, vectorized implementation.
* A more complex implementation based on rolling means.
These implementations are compared to one another and *stats::density* with respect to their estimates and runtime.

## plot_density
The *plot_density* density function serves as a graphical tool for determining if density estimates are conform with corresponding histogram representations. We perform no formal testing of this.

# TODO
* Implement CV stopping for *iter_dens_est*.
* Write some benchmarks for the *iter_bw_est* function.
* Write the alternative *compile_density* methods.










